Adaptive Sequential Model Selection

### 2015-05-05 paper.tex
Filled in most of "sub-path sufficiency principle" section

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\tV}{\widetilde{V}}
\newcommand{\tR}{\widetilde{R}}
\newcommand{\hK}{\widehat{K}}
\newcommand{\hk}{\hat{k}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}

@@

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Mon May  4 21:11:15 2015
\begin{table}[ht]
  \centering
  \begin{tabular}{l|ccccc|c}
%    & & & $p_2$ & & & \\
    \multicolumn{7}{c}{$p_2(Y)$}\\[5pt]
    ${\large p_1(Y)}$ & (0,0.2] & (0.2,0.4] & (0.4,0.6] & (0.6,0.8] & (0.8,1] & \textbf{Total} \\ 
    \hline
    (0,0.2] & 1.0 & 2.7 & 4.2 & 5.6 & 6.7 & 20.1 \\ 
    (0.2,0.4] & 1.4 & 3.4 & 4.5 & 5.2 & 5.5 & 20.0 \\ 
    (0.4,0.6] & 2.3 & 4.3 & 4.7 & 4.5 & 4.2 & 20.0 \\ 
    (0.6,0.8] & 4.2 & 5.4 & 4.3 & 3.4 & 2.7 & 20.0 \\ 
    (0.8,1] & 11.1 & 4.3 & 2.3 & 1.4 & 1.0 & 20.0 \\ 
    \hline
    \textbf{Total} & 19.9 & 20.0 & 20.0 & 20.1 & 20.0 & 100.0 \\ 
  \end{tabular}
  \caption{Two-way contingency table of saturated-model $p$-values $(p_1(Y), p_2(Y))$ for Example~\ref{ex:bivariate}, after binning into cells of height and width 0.2. We report the percentage of $p$-value pairs falling into each cell out of one million simulations from the global null hypothesis, $\mu=0$. Both $p$-values are marginally uniform but strongly dependent, with a correlation of $-48\%$.}
\label{tab:bv_twoWayTable}
\end{table}

@@

A $p$-value for testing model $M_{k-1}$ against $M_k$ satisfies the sufficiency principle if it is a function only of the sufficient statistics of $M_k$

@@

We say that a selection algorithm $M_{[d]}(\cdot)$ satisfies the {\em sub-path sufficiency principle} (henceforth SPSP) if, given $M_k(Y)=m_k$, the sequence of models $M_{[k]}$ depends only on sufficient statistics of $m_k$. That is

A model selection procedure derived from the ``ever-active'' coefficients of a regularized likelihood path always satisfies the SPSP. Let $M_\infty$ be a model with parameters $\theta\in\Theta\sub \R^p$, and likelihood $L(\theta; Y)$. For $r=0,\ldots,R$, let $P_r(\theta)$ denote a sequence of regularizing penalties, and define
\begin{align}\label{eq:regPathDef_start}
  \hat\theta^{r}(Y) &\triangleq 
  \argmin_{\theta\in\Theta} -\log L(\theta; Y) + P_r(\theta) \\
  \tE_r(Y) &\triangleq \left\{j:\; \hat\theta_j^s \neq 0 
    \text{ for any } s \leq r \right\}
\end{align}
Note that while the sets $\tE_r$ are nested by definition, we could have $\tE_r = \tE_{r+1}$ for most values of $r$. Let $r(0)=0$, and for $k>0$ define
\begin{equation}
r(k) = \min\{s:\; \tE_s \neq \tE_{r(k-1)}\}.
\end{equation}
That is, the $r(k)$ are the indices where the active set changes. Let $E_k = \tE_{r(k)}$, and take
\begin{equation}\label{eq:regPathDef_end}
M_k \triangleq \{F\in M_\infty:\; \theta_j = 0,\; \forall j \notin E_k\}
\end{equation}

Model paths described in this way satisfy the SPSP, as we see next.

\begin{proposition}[Regularized Likelihood Paths Satisfy SPSP]
  Model paths specified as in (\ref{eq:regPathDef_start}--\ref{eq:regPathDef_end}) satisfy the sub-path sufficiency principle.
\end{proposition}
\WFcomment{Actually, you just keep the times when the model changes... Need to recast notation a bit; indices of $\lambda$ (or $P$, here) need not match up with indices of model.}

\begin{proof}
  Let 
\end{proof}

\begin{proposition}[Forward Stepwise Paths Satisfy SPSP]
  \WFcomment{Generalized forward stepwise}
\end{proposition}

\begin{proof}
  Let
\end{proof}

@@

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Fri May  1 09:29:28 2015
\begin{table}[ht]
  \centering
  \begin{tabular}{llcccc}
    \hline
    & & $p_{\text{screen}}$ & $\text{FWER}_{\text{mod}}$ 
    & $\text{FDR}_{\text{mod}}$ 
    & $\text{FDR}_{\text{var}}$ \\ 
    \hline
    Selected & Simple & .290 & .000 & .002 & .039 \\ 
    Selected & Forward & .559 & .027 & .020 & .066 \\ 
    Selected & Strong & .041 & .014 & .008 & .042 \\ 
    \hline
    Saturated & Simple & .000 & .000 & .000 & .028 \\ 
    Saturated & Forward & .014 & .000 & .000 & .030 \\ 
    Saturated & Strong & .000 & .000 & .000 & .032 \\ 
    \hline
    Knockoffs & & .000 &  &  & .231 \\ 
    \hline
  \end{tabular}
  \caption{\WFcomment{Comment on results once they are correct. These numbers are wrong right now.}}
\end{table}

>>>>

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\tV}{\widetilde{V}}
\newcommand{\tR}{\widetilde{R}}
\newcommand{\tL}{\widetilde{L}}
\newcommand{\hk}{\hat{k}}
\newcommand{\hr}{\hat{r}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}

@@

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Mon May  4 21:11:15 2015
\begin{table}[ht]
  \centering
  \begin{tabular}{l|ccccc|c}
    \multicolumn{7}{c}{Saturated-Model $p$-Values 
      (\% of $10^6$ Simulations)}\\
    \hline
    \multicolumn{7}{c}{$p_2(Y)$}\\[5pt]
    ${\large p_1(Y)}$ & (0,0.2] & (0.2,0.4] & (0.4,0.6] & (0.6,0.8] & (0.8,1] & \textbf{Total} \\ 
    \hline
    (0,0.2] & 1.0 & 2.7 & 4.2 & 5.6 & 6.7 & 20.1 \\ 
    (0.2,0.4] & 1.4 & 3.4 & 4.5 & 5.2 & 5.5 & 20.0 \\ 
    (0.4,0.6] & 2.3 & 4.3 & 4.7 & 4.5 & 4.2 & 20.0 \\ 
    (0.6,0.8] & 4.2 & 5.4 & 4.3 & 3.4 & 2.7 & 20.0 \\ 
    (0.8,1] & 11.1 & 4.3 & 2.3 & 1.4 & 1.0 & 20.0 \\ 
    \hline
    \textbf{Total} & 19.9 & 20.0 & 20.0 & 20.1 & 20.0 & 100.0 \\ 
    \hline
  \end{tabular}
  \caption{Two-way contingency table of saturated-model $p$-values $(p_1(Y), p_2(Y))$ for Example~\ref{ex:bivariate}, after binning into cells of height and width 0.2. We report the percentage of $p$-value pairs falling into each cell out of one million simulations from the global null hypothesis, $\mu=0$. Both $p$-values are marginally uniform but strongly dependent, with a correlation of $-48\%$.}
\label{tab:bv_twoWayTable}
\end{table}

@@

A $p$-value for testing model $M_{k-1}$ against $M_k$ satisfies the (conditional) {\em sufficiency principle} if, on the event $\{M_{k-1}=m_{k-1}, M_k=m_k\}$, it depends only on sufficient statistics of $M_k$.

Based on the law
\[
\L_{m_{k-1}}\left(T_k \;\;\mid\;\; T_{k-1},\; M_{k-1}=m_{k-1},\; M_k=m_k\right)
\]

If $T(Y)$ is a complete sufficient statistic, then any exact $p$-value must have the latter property.

@@

We say that a selection algorithm $M_{[d]}(\cdot)$ satisfies the {\em sub-path sufficiency principle} (henceforth SPSP) if, on the event $M_k(Y)=m_k$, the sequence of models $M_{[k]}$ depends only on sufficient statistics of $m_k$. Again, if

It will be useful to introduce the following notation:

\begin{model}[Generalized Sparse Model]\label{mod:genSparse}
  Let $M$ be a model parameterized by $\theta\in \Theta \sub \R^p$:
  \[
  M = \{F_\theta:\; \theta \in \Theta\}.
  \]
  For subsets $E\sub [p]$ define the submodel with active coefficients $E$ as follows:
  \[
  \Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
  \quad M(E) = \{F_\theta:\; \theta\in \Theta_E\}.
  \]
  Let $L(\theta; Y)$ denote the likelihood for model $M$.
\end{model}
Note that, although we will make no assumptions about the form of $M$, the results of this section are only useful in cases where we can actually construct single-step $p$-values. At present, that caveat largely restricts us to using exponential family models.

In Model~\ref{mod:genSparse}, the {\em forward stepwise} algorithm proceeds as follows: we first set $E_0(Y)$ arbitrarily, then at step $k=1,\ldots,d$, we define
\begin{align}\label{eq:forwardDef_start}
j_k &= \argmax_j \;\;\sup \left\{L(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\} \\
E_k &= E_{k-1} \cup \{j_k\}\\\label{eq:forwardDef_end}
M_k &= M(E_k)
\end{align}

To make this definition more concrete, we can specialize it to several familiar examples:

\begin{example}[Forward Stepwise Regression]
In the case of linear regression with known $\sigma^2$, $p$ is the number of variables and $M$ is $\cN(X\beta, \;\sigma^2I)$. For active set $E \sub [p]$, $M(E)$ is the model $\cN(X_E\beta_E, \;\sigma^2I)$, and $E_0 = \emptyset$. Because the maximized log-likelihood for a linear regression is monotone in the residual sum of squares, the forward stepwise algorithm adds whichever variable reduces the residual sum of squares the most.

For linear regression with unknown $\sigma^2$, there are $p+1$ parameters $(\beta, \sigma^2)$ and $E_0 = \{p+1\}$ because $\sigma^2$ is a parameter for every model.
\end{example}

\begin{example}[Principle Components Analysis]
\WFcomment{Do this one.}  
\end{example}

All forward stepwise procedures satisfy the SPSP.

\begin{proposition}[Forward Stepwise Satisfies SPSP]\label{prop:forwardSPSP}
  The forward stepwise algorithm in Model~\ref{mod:genSparse} as defined in (\ref{eq:forwardDef_start}--\ref{eq:forwardDef_end}) satisfies the SPSP.
\end{proposition}
\begin{proof}
  Let $A$ denote the event $\{M_k = M(E)\}$ for some fixed active set $E$. Letting $T(Y)$ be a sufficient statistic for $m_k$, the likelihood $L$ restricted to $\Theta(E) \times A$ is proportional to a function depending only on $T$. If $L_T$ is the likelihood of $T(Y)$, then
\[
L(\theta; y) = L_T(\theta; T(y)) \;L_{Y\mid T}(y), \;\; \text{ for } (\theta, y) \in  \Theta(E) \times A.
\]
Note that $L_{Y \mid T}(y)$ does not depend on $\theta$ since $T$ is sufficient for $M(E)$.

For $y\in A$ and $\ell \leq k$, we have $E_\ell(y) \sub E$. This means that the maximizer in \eqref{eq:forwardDef_start} is attained for some $j\in E$. So, for $\ell \leq k$, on $A$,
\begin{align}
  j_\ell &= \argmax_{j\in E} \;\;\sup \left\{L(\theta; Y):\;
    \theta\in\Theta(E_{\ell-1} \cup \{j\})\right\} \\
  &= \argmax_{j\in E} \;\;\sup \left\{L_T(\theta; T(Y)):\;
    \theta\in\Theta(E_{\ell-1} \cup \{j\})\right\}. \label{eq:onlyT}
\end{align}
Replacing $L$ with $L_T$ in~\eqref{eq:onlyT} is justified because $\Theta(E_{\ell - 1} \cup \{j\}) \sub \Theta(E)$ when $E_{\ell-1}\cup E$ and $j\in E$.

Equation~\eqref{eq:onlyT} shows that $j_1,\ldots, j_{k-1}$ all depend on $Y$ only through $T(Y)$. As a result, it also follows that the sequence $M_{[k-1]}(Y)$ depends only of $T(Y)$.
\end{proof}

Another class of model selection procedures satisfying the SPSP is the sequence ``ever-active'' sets in regularized likelihood methods. For $r=0,1,\ldots$, let $P_r(\theta)$ denote a regularizing penalty, and define
\begin{align}\label{eq:regPathDef_start}
  \hat\theta^{r}(Y) &= 
  \argmin_{\theta\in\Theta} -\log L(\theta; Y) + P_r(\theta) \\
  \tE_r(Y) &= \left\{j:\; \hat\theta_j^s \neq 0 
    \text{ for any } s \leq r \right\}
\end{align}

While the sets $\tE_r$ are nested by definition, we could have $\tE_r = \tE_{r+1}$ for most values of $r$. We will take the sequence of {\em distinct} ever-active sets. Let $R_0=0$, and for $k>0$ let $R_k$ denote the (random) index where the active set actually changes:
\begin{align}
  R_k &= \min\{s:\; \tE_s \neq \tE_{R_{k-1}}\}\\
  E_k &= \tE_{R_k}\\
  \label{eq:regPathDef_end}
  M_k &= M(E_k)
\end{align}
\WFcomment{Slightly sloppy definitions: note that $d$, the number of models, is random. $d$ could be 0 or could be infinite, depending on the data. Also, it isn't really necessary that the $r$ values be integers...}

\begin{proposition}[Regularized Likelihood Paths Satisfy SPSP]\label{prop:regPathSPSP}
Ever-active model paths derived from regularized likelihood methods as in (\ref{eq:regPathDef_start}--\ref{eq:regPathDef_end}) satisfy the sub-path sufficiency principle.
\end{proposition}

\begin{proof}
  Again, let $A$ denote the event $\{M_k = M(E)\}$ for some fixed $E$, and define $T$ and $L_T$ as in the proof of Proposition~\ref{prop:forwardSPSP}. We can partition $A$ into subsets $A_r = A \cap \{R(k)=r\}$.

On the event $A_r$, every $\hat\theta^s$ for $s\leq r$ is supported on $E$. That is, for $s \leq r$, we can write 
\begin{equation}\label{eq:onlyT_Ar}
  \hat\theta^{s} = 
  \argmin_{\theta\in\Theta(E)} -\log L_T(\theta; T(Y)) + P_s(\theta) \quad \text{ on } A_r
\end{equation}
Equation~\label{eq:onlyT_Ar} guarantees that $M_{[k-1]}$ is a function of $T(Y)$ on $A_r$. That is, we can write
\[
M_{[k-1]}(Y) = M_{[k-1]}^r(T(Y)) \quad \text{ on } A_r
\]
It remains only to show that, on $A$, $R_k(Y)$ depends only on $T(Y)$; then we have
\[
M_{[k-1]}(Y) = M_{[k-1]}^{R(T)}(T) \quad \text{ on } A
\]
But $R_k$ is a function of $T$ on $A$ for the same reason: that $\hat\theta^s$ depends only on $T$ until other coefficients enter the model.
\WFcomment{Would the proof be cleaner with $A_r$ defined as $A\cap \{R_k \leq r\}$?}
\end{proof}

@@

\WFcomment{Should we compute the FWER and FDR only conditional on screening? Strong / Forward / Simple should all be correct given screening since they condition on $p_{[k_0-1]}$.}

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Fri May  1 09:29:28 2015
\begin{table}[ht]
  \centering
  \begin{tabular}{llcccc}
    \hline
    Method & Stopping Rule & $p_{\text{screen}}$ & $\text{FWER}_{\text{mod}}$ 
    & $\text{FDR}_{\text{mod}}$ 
    & $\text{FDR}_{\text{var}}$ \\ 
    \hline
    Selected & Basic & .290 & .000 & .002 & .039 \\ 
    Selected & Forward & .559 & .027 & .020 & .066 \\ 
    Selected & Strong & .041 & .014 & .008 & .042 \\ 
    \hline
    Saturated & Basic & .000 & .000 & .000 & .028 \\ 
    Saturated & Forward & .014 & .000 & .000 & .030 \\ 
    Saturated & Strong & .000 & .000 & .000 & .032 \\ 
    \hline
    Knockoffs & & .000 & --- & --- & .231 \\ 
    \hline
  \end{tabular}
  \caption{\WFcomment{Comment on results once they are correct. These numbers are wrong right now.}}
\end{table}
